
---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# innsight - Get the insights of your Neural Network

<!-- badges: start -->
<!-- badges: end -->

`innsight` is an R package that interprets the behavior and explains predictions
of machine learning models. It implements several model-specific interpretability
methods based on Neural Networks in R, e.g., 

* Layer-wise Relevance Propagation ([LRP](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0130140))
* Deep Learning Important Features ([DeepLift](https://arxiv.org/abs/1704.02685))
* Gradient-based methods:
  * Vanilla Gradient, including [Gradient x Input](https://www.jmlr.org/papers/v11/baehrens10a.html)
  * Smoothed gradients ([SmoothGrad](https://arxiv.org/abs/1706.03825))
* Connection Weights

Basically, a Neural Network of the libraries [`torch`](https://torch.mlverse.org/),
[`keras`](https://keras.rstudio.com/) and [`neuralnet`](https://cran.r-project.org/web/packages/neuralnet/)
can be passed, which is internally converted into a torch model with special
insights needed for interpretation.

## Installation

Install `devtools` and `torch` from CRAN. Than run the following commands:
```{r, echo=FALSE}
library(devtools)
library(torch)
```

```{r, eval = FALSE}
devtools::install_github("bips-hb/innsight")
library(innsight)

```

## Examples

### 1. Neuralnet (Simple)

Let's train a simple fully connected Neural Network with the package `neuralnet`
on the iris dataset:

```{r, echo=TRUE}
library(innsight)
library(neuralnet)
data(iris)

nn <- neuralnet( Species ~ .,
                 iris, linear.output = FALSE,
                 hidden = c(10,8), act.fct = "tanh", threshold = 0.5)
# create an analyzer for this model
analyzer = Analyzer$new(nn)

# Apply method Connection_Weights
cw <- Connection_Weights$new(analyzer)
cw$get_result()
```

### 2. keras (advanced)

```{r}
library(keras)

fashion_mnist <- dataset_fashion_mnist()
c(train_images, train_labels) %<-% fashion_mnist$train
c(test_images, test_labels) %<-% fashion_mnist$test


train_images <- train_images / 255
test_images <- train_images[1:100,,] / 255
dim(train_images) <- c(60000, 28, 28, 1)
dim(test_images) <- c(100, 28, 28, 1)

model <- keras_model_sequential()
model %>%
  layer_conv_2d(filters = 8, kernel_size = c(4,4), input_shape = c(28,28, 1), activation = 'relu') %>%
  layer_conv_2d(filters = 4, kernel_size = c(2,2), activation = 'relu') %>%
  layer_flatten() %>%
  layer_dropout(0.25) %>%
  layer_dense(units = 128, activation = 'relu') %>%
  layer_dense(units = 10, activation = 'softmax')

model %>% compile(
  optimizer = 'adam', 
  loss = 'sparse_categorical_crossentropy',
  metrics = c('accuracy')
)

model %>% fit(train_images, train_labels, epochs = 10, verbose = 2)

analyzer <- Analyzer$new(model)

lrp <- LRP$new(analyzer, test_images, channels_first = FALSE)
result <- lrp$get_result()
```
```{r, echo = FALSE}
library(tidyr)
library(ggplot2)
library(scales)
library(gridExtra)

class_names = c('T-shirt/top',
                'Trouser',
                'Pullover',
                'Dress',
                'Coat', 
                'Sandal',
                'Shirt',
                'Sneaker',
                'Bag',
                'Ankle boot')


img <- test_images[1,,,]
image_1 <- as.data.frame(img)
colnames(image_1) <- seq_len(ncol(image_1))
image_1$y <- seq_len(nrow(image_1))
image_1 <- gather(image_1, "x", "value", -y)
image_1$x <- as.integer(image_1$x)

g1 <- ggplot(image_1, aes(x = x, y = y, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "black", na.value = NA) +
  scale_y_reverse() +
  theme_minimal() +
  theme(panel.grid = element_blank())   +
  theme(aspect.ratio = 1) +
  xlab("") +
  ylab("")

img <- result[1,,,1, test_labels[1]]
image_1 <- as.data.frame(img)
colnames(image_1) <- seq_len(ncol(image_1))
image_1$y <- seq_len(nrow(image_1))
image_1 <- gather(image_1, "x", "value", -y)
image_1$x <- as.integer(image_1$x)

g2 <- ggplot(image_1, aes(x = x, y = y, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = muted("blue"),mid = "white", high = muted("red"), na.value = NA) +
  scale_y_reverse() +
  theme_minimal() +
  theme(panel.grid = element_blank())   +
  theme(aspect.ratio = 1) +
  xlab("") +
  ylab("")

grid.arrange(g1, g2, ncol=2)
```

