---
title: "Example 1: Iris Dataset with `torch`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
always_allow_html: yes
vignette: >
  %\VignetteIndexEntry{Example 1: Iris Dataset with `torch`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  size = "huge",
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  out.width = "95%"
)
```

```{r, echo = FALSE}
Sys.setenv(LANG = "en_US.UTF-8")
set.seed(1111)
options(width = 500)
torch::torch_manual_seed(1111)
```

In this very simple example, we want to show you the use of the **innsight** 
package and its possibilities in the simplest case. For this, we use the Iris 
dataset, which contains the length and width of the sepals and petals of the 
three flower species Iris setosa, versicolor and virginica. In total, 
there are measurements of $50$ flowers of each type in centimeters.

## Step 0: Train a model

The first step does not necessarily need the **innsight** package yet, but at 
this point, it is listed anyway to have a fully working example. In this 
example, the model is created in **torch** and trained as well.

**Note:** It is also possible to train the **torch** model using the 
[`luz`](https://mlverse.github.io/luz/) package, which is described in 
the [vignette with the penguin dataset](https://bips-hb.github.io/innsight/articles/Example_2_penguin.html) (see `?vignette(package='innsight'`)).

```{r example_1_train, echo = TRUE, eval = TRUE}
library(innsight)
library(torch)

# Set seeds for reproducibility
set.seed(42)
torch_manual_seed(42)

# Prepare Data
data(iris)
x <- torch_tensor(as.matrix(iris[, -5]))
# normalize input to [-1, 1]
x <- 2 * (x - x$min(1)[[1]]) / (x$max(1)[[1]] - x$min(1)[[1]]) - 1
y <- torch_tensor(as.integer(iris[, 5]))


# normalize input data

# Define Model
model <- nn_sequential(
  nn_linear(4, 30),
  nn_relu(),
  nn_dropout(0.3),
  nn_linear(30, 10),
  nn_relu(),
  nn_linear(10, 3),
  nn_softmax(dim = 2)
)

# Train model
optimizer <- optim_adam(model$parameters, lr = 0.001)
for (t in 1:2500) {
  y_pred <- torch_log(model(x))
  loss <- nnf_nll_loss(y_pred, y)
  if (t %% 250 == 0) {
    cat("Loss: ", as.numeric(loss), "\n")
  }
  optimizer$zero_grad()
  loss$backward()
  optimizer$step()
}
```

## Step 1: Convert the model

The easiest way to convert a model for the **innsight** package is just using the 
model for the Converter class without any further arguments (except for **torch** 
models the input dimension is still needed). Thus, the converting part for our 
trained model is done by:

```{r example_1_conv_1}
converter <- Converter$new(model, input_dim = c(4))
```

Since there are no labels for the inputs and outputs in the **torch** model, they 
were generated automatically by **innsight**. But you can pass them by yourself
using the corresponding arguments. These labels are then also used in the 
visualizations in [step 4](#step-3-visualization).

```{r example_1_conv_2}
converter <- Converter$new(model,
  input_dim = c(4),
  input_names = c("Sepal (length)", "Sepal (width)", "Petal (length)", "Petal (width)"),
  output_names = c("Setosa", "Versicolor", "Virginica")
)
```

## Step 2: Apply methods

Now we can apply the implemented methods to our trained model. Although these 
methods are different, they are called more or less the same way in 
**innsight**. Essential arguments are of course the converter object 
(`converter`) and the data (`data`) to which the method is to be applied.
In addition to the essential and method-specific arguments, the following other 
arguments can be set for each (local) method:

- `channels_first`: There are, basically, two data formats for image and 
signal data, having the channels at the last position or directly after the 
batch axis. To be able to handle both formats, the format of the passed data 
can be specified with this argument. In this example it is unused because we 
have tabular data.

- `output_idx`: These indices specify the output nodes (or classes) for 
which the method is to be applied. By default, the result for the 
first ten classes is calculated for efficiency reasons. However, if you want to 
examine several specific classes at once, you must specify the corresponding 
indices with this argument.

- `ignore_last_act`: Sometimes it makes more sense to ignore the last 
activation function to get better results. This is especially true for 
classification problems with softmax or sigmoid as the last activation.

For more and detailed information on the arguments see [this article](https://bips-hb.github.io/innsight/vignette/detailed_overview.html#step-2-apply-selected-method). In the following, we apply our trained model to different 
methods provided in the package:

<ul>
<li>
**Gradient without last activation**
```{r, message = FALSE, results = 'hide'}
grad_no_softmax <- Gradient$new(converter, x, ignore_last_act = TRUE)
```
</li>
<li>
**Gradient with last activation**
```{r, message = FALSE, results = 'hide'}
grad_softmax <- Gradient$new(converter, x, ignore_last_act = FALSE)
```
</li>
<li>
**LRP with $\varepsilon$-rule ($\varepsilon = 0.01$)**

We can also set the method-specific arguments, such as the LRP rule and its 
associated parameter:

```{r, message = FALSE, results = 'hide'}
lrp_eps <- LRP$new(converter, x, rule_name = "epsilon", rule_param = 0.01)
```
</li>
<li>
**DeepLift with mean as reference value**

```{r, message = FALSE, results = 'hide'}
x_ref <- x$mean(1, keepdim = TRUE) # ref value need the shape (1,4)
deeplift_mean <- DeepLift$new(converter, x, x_ref = x_ref)
```
</li>
</ul>

Afterwards the results can be retrieved either with the class method 
`method$get_result` or the equivalent S3 function `get_result`. With the 
argument `type`, it is also possible to select whether the results are returned 
as `array`, `data.frame` or `torch_tensor`:

```{r}
# get result as a data.frame with the class method
head(grad_no_softmax$get_result(type = "data.frame"), 5)

# get result as array (default) with the generic S3 function
str(get_result(grad_no_softmax))
```

## Step 3: Visualization

The package **innsight** provides two ways to visualize the results of a method, 
namely as `innsight_ggplot2` or `innsight_plotly` object. Both are S4 classes 
to combine multiple plots nicely and to be able to make visual modifications 
or adjustments to the selection of plots even after the object has been created.
The first class is based on [**ggplot2**](https://ggplot2.tidyverse.org/) and
behaves partly like an ordinary **ggplot2** object. Whereas the other one is 
based on the [**plotly**](https://plotly.com/r/) package and creates an 
interactive graph with more detailed information about each variable.
For more information on the S4 classes `innsight_ggplot`
and `innsight_plotly` see the [in-depth vignette](https://bips-hb.github.io/innsight/vignette/detailed_overview.html#advanced-plotting) or the respective R documentation (`?innsight_ggplot2` or `?innsight_plotly`).

For each of these classes and thus of course also for each method, there are 
two plot functions, `plot` shows only individual data points and 
`boxplot` visualizes summaries of multiple data points using summary statistics.

#### Plot individual results

The function `plot` is implemented for each of the available methods. You can 
select your desired data points and output nodes/classes with the `data_idx` 
and `output_idx` arguments, respectively. To switch between a **ggplot2** and 
**plotly`** based plot, you can use the logical `as_plotly` parameter, but 
this requires a successful installation of this package.

**Note:** Of course, only results of output nodes/classes to which the method 
has been applied can be plotted, i.e. the indices in the `output_idx` 
argument must be a subset of the argument with the same name when calling 
the method in Step 2.

**ggplot2-based plot**

Despite being an object of the class `innsight_ggplot2`, it can be treated 
like a normal **ggplot2** object, i.e. geoms, scales and themes can be added 
and modified.

```{r, fig.height=6, fig.keep='all', fig.width=9}
# Show data point 1 and 111 for output node 1 (Setosa) and 2 (Versicolor)
plot(grad_no_softmax, data_idx = c(1, 111), output_idx = c(1, 2)) +
  ggplot2::theme_bw()
```

**plotly-based plot**

```{r, fig.height=4, fig.keep='all', fig.width=9, eval = FALSE}
# Show data point 1 for output node 1 (Setosa) and 2 (Versicolor)
plot(deeplift_mean, data_idx = 1, output_idx = c(1, 2), as_plotly = TRUE)
```
```{r, fig.height=4, echo = FALSE, fig.width=9, message = FALSE, eval = Sys.getenv("RENDER_PLOTLY", unset = 0) == 1}
# Show data point 1 for output node 1 (Setosa) and 2 (Versicolor)
p <- plot(deeplift_mean, data_idx = 1, output_idx = c(1, 2), as_plotly = TRUE)
plotly::config(print(p, shareY = TRUE))
```
```{r, eval=Sys.getenv("RENDER_PLOTLY", unset = 0) == 0, echo = FALSE}
warning("Due to the size, this image was not generated in the R vignette! See the innsight webpage for this image: https://bips-hb.github.io/innsight/articles/Example_1_iris.html#plot-individual-results")
```

#### Plot summarized results

The S3 function `boxplot` is implemented for each of the available local methods. 
You can select your desired data points (default is `'all'`) and output 
nodes/classes with the `data_idx` and `output_idx` arguments, respectively. 
To switch between a **ggplot2** and **plotly** based plot, you can use the 
logical `as_plotly` parameter, but this requires a successful installation of 
this package. In addition, you can use `ref_data_idx` to select a single data 
point that will be visualized in red as a reference value, and 
`preprocess_FUN` to select a function that will be applied to all data 
in advance (e.g. the absolute value).

**Note:** Of course, only results of output nodes/classes to which the method 
has been applied can be plotted, i.e. the indices in the `output_idx` 
argument must be subset of the argument with the same name when calling 
the method in Step 2.

**ggplot2-based plot**

Despite being an object of the class `innsight_ggplot2`, it can be treated 
like a normal **ggplot2** object, i.e. geoms, scales and themes can be added 
and modified.

```{r, fig.height=6, fig.keep='all', fig.width=9}
# Summarized results for output node 1 (Setosa) and 2 (Versicolor) and
# reference value 3
boxplot(grad_no_softmax, output_idx = c(1, 2), ref_data_idx = 3, preprocess_FUN = abs) +
  ggplot2::theme_bw()
```

**plotly-based plot**

```{r, fig.height=4, fig.keep='all', fig.width=9, eval = FALSE}
# Show boxplot only for instances of class setosa for output node 1 (Setosa)
# and 2 (Versicolor)
boxplot(lrp_eps, data_idx = 1:50, output_idx = c(1, 2), as_plotly = TRUE)
```
```{r, fig.height=4, echo = FALSE, fig.width=9, message = FALSE, eval = Sys.getenv("RENDER_PLOTLY", unset = 0) == 1}
# Show data point 1 for output node 1 (Setosa) and 2 (Versicolor)
p <- boxplot(lrp_eps, data_idx = 1:50, output_idx = c(1, 2), as_plotly = TRUE)
plotly::config(print(p, shareY = TRUE))
```
```{r, eval=Sys.getenv("RENDER_PLOTLY", unset = 0) == 0, echo = FALSE}
warning("Due to the size, this image was not generated in the R vignette! See the innsight webpage for this image: https://bips-hb.github.io/innsight/articles/Example_1_iris.html#plot-summarized-results")
```
