---
title: "Introduction to `innsight`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to `innsight`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```


In the last decade, there has been a tremendous rush and growth in machine 
learning. Almost every year, new architectures break records on scientific test
data sets, and the number of layers continues to grow through regularization 
methods making today's neural networks far more complex than an original 
linear or logistic model. Nevertheless, the scientific focus is more on the 
predictive power than on the interpretability and the respective methods that 
already exist for the interpretation of single predictions or the whole 
neural networks are only sparsely or rarely implemented in the R programming
language. The **`innsight`** package addresses this lack of interpretable 
machine learning methods in R (similar to 
[**`iNNvestigate`**](https://github.com/albermax/innvestigate) for Python) and 
provides the most popular methods for neural networks trained in R for 
analysis in one unified interface.

The steps for explaining individual predictions with the provided methods 
are unified in this package and follow a strict scheme. This will hopefully 
allow any user a smooth and easy introduction to the possibilities of 
this package. The steps are:

```{r, eval = FALSE}
# Step 1: Model creation and converting
model = ...
converter <- Converter$new(model)

# Step 2: Apply selected method to your data
result <- Method$new(converter, data)

# Step 3: Plot the results
plot(result) 
boxplot(result)
```

## Step 1: Model Creation and Converting
The `innsight` package aims to be as flexible as possible and independent of 
any particular deep learning package in which the passed network was learned 
or defined. For this reason, there are several ways in this package to pass 
a neural network and then interpret their predictions.

### Library `torch`
Currently, only models created by `torch::nn_sequential` are accepted. 
However, the most popular standard layers and activation functions are 
available:

* **Linear layers:** `nn_linear`
* **Convolutional layers:** `nn_conv1d`, `nn_conv2d` (but only with 
`pading_mode = "zeros"` and numerical `padding`)
* **Max-pooling layers:** `nn_max_pool1d`, `nn_max_pool2d` (both only with default 
arguments for `padding` (`0`), `dilation` (`1`), `return_indices` (`FALSE)`) 
and `ceil_mode` (`FALSE`))
* **Average-pooling layers:** `nn_avg_pool1d`, `nn_avg_pool2d` (both only with 
default arguments for `padding` (`0`), `return_indices` (`FALSE)`) 
and `ceil_mode` (`FALSE`))
* **Flatten layer:** `nn_flatten` (see [torch issue #716](https://github.com/mlverse/torch/issues/716#issuecomment-946117545) and use
`classname = "nn_flatten"`)
* **Skipped layers:** `nn_dropout`
* **Activation functions:** `nn_relu`, `nn_leaky_relu`, `nn_softplus`,
`nn_sigmoid`, `nn_softmax`, `nn_tanh` (contact me if you need any more)

If you want to create an instance of the class `Converter` with a torch 
model that meets the above conditions, you have to keep the following things 
in mind:

* In a torch model, the shape of the inputs is not stored; hence it must be
specified with argument `input_dim` within the initialization of `Converter`.

* If no further arguments are set for the `Converter` instance, default labels
are generated for the input (`'X1'`, `'X2'`, ...) and output names
(`'Y1'`, `'Y2'`, ... ). In torch models this information is not stored 
or often it is not even present in the training data. In the converter, 
however, there is the possibility (argument `input_names` and `output_names`) 
to pass the names, which will then be used in all results and plots 
created by this object.

#### Example 1: Dense Network
```{r}
library(torch)
library(innsight)

# Create model
model <- nn_sequential(
  nn_linear(3, 10),
  nn_relu(),
  nn_linear(10, 2, bias = FALSE),
  nn_softmax(2)
)
# Convert the model
conv_dense <- Converter$new(model, input_dim = c(3))
# Convert model with input and output names
conv_dense_with_names <- 
  Converter$new(model, input_dim = c(3),
                input_names = list(c("Price", "Weight", "Height")),
                output_names = list(c("Buy it!", "Don't buy it!")))
# Show output names
conv_dense_with_names$model_dict$output_names
```

#### Example 2: CNN
Unfortunately, at the moment (`torch = 0.6.0`) it is not possible to use a flatten module 
in `nn_sequential`, because it is only implemented as a function 
(`torch_flatten`). For this reason we define our own module as described in
[torch issue #716](https://github.com/mlverse/torch/issues/716#issuecomment-946117545).
To ensure that our package recognizes this module as a flattening layer, 
`classname = "nn_flatten"` must be set.
```{r}
nn_flatten <- nn_module(
    classname = "nn_flatten",
    initialize = function(start_dim = 2, end_dim = -1) {
      self$start_dim <- start_dim
      self$end_dim <- end_dim
    },
    forward = function(x) {
      torch_flatten(x, start_dim = self$start_dim, end_dim = self$end_dim)
    }
  )
```

```{r}
# Create CNN for images of size (3, 28, 28)
model <- nn_sequential(
  nn_conv2d(3, 5, c(2, 2)),
  nn_relu(),
  nn_max_pool2d(c(1,2)),
  nn_conv2d(5, 6, c(2, 3), stride = c(1, 2)),
  nn_relu(),
  nn_conv2d(6, 2, c(2, 2), dilation = c(1, 2), padding = c(5,4)),
  nn_relu(),
  nn_avg_pool2d(c(2,2)),
  nn_flatten(),
  nn_linear(48, 5),
  nn_softmax(2)
)

# Convert the model
conv_cnn <- Converter$new(model, input_dim = c(3, 10, 10))
```

### Library `keras`
Keras models created by `keras_model_sequential` or `keras_model` are 
accepted. Within these functions, the following layers are allowed to be used:

* **Linear layers:** `layer_dense`
* **Convolutional layers:** `layer_conv_1d`, `layer_conv_2d`
* **Max-pooling layers:** `layer_max_pooling_1d`, `layer_max_pooling_2d`
* **Average-pooling layers:** `layer_average_pooling_1d`, `layer_average_pooling_2d`
* **Flatten layer:** `layer_flatten`
* **Skipped layers:** `layer_dropout`
* **Activation functions:** Activation functions are currently only allowed 
as a string argument in the individual layers (`activation`) and not as a 
standalone layer like e.g. `layer_activation_relu`. The following are accepted
as arguments: `"relu"`, `"softplus"`, `"sigmoid"`,
`"softmax"`, `"tanh"`, `"linear"` (contact me if you need any more)

Analogous to a torch model, no names of inputs or outputs are stored in 
a keras model, i.e. if no further arguments are set for the `Converter`
instance, default labels are generated for the input (e.g. `'X1'`, `'X2'`, ...) 
and output names (`'Y1'`, `'Y2'`, ... ). In the converter, 
however, there is the possibility (argument `input_names` and `output_names`) 
to pass the names, which will then be used in all results and plots 
created by this object.

#### Example 1: Dense Network
```{r}
library(keras)

# Create model
model <- keras_model_sequential()
model <- model %>%
  layer_dense(10, input_shape = c(5), activation = "softplus") %>%
  layer_dense(8, use_bias = FALSE, activation = "tanh") %>%
  layer_dropout(0.2) %>%
  layer_dense(4, activation = "softmax")

# Convert the model
conv_dense <- Converter$new(model)
```

#### Example 2: CNN
```{r}
library(keras)

# Create model
model <- keras_model_sequential()
model <- model %>%
  layer_conv_2d(4, c(5,4), input_shape = c(10,10,3), activation = "softplus") %>%
  layer_max_pooling_2d(c(2,2), strides = c(1,1)) %>%
  layer_conv_2d(6, c(3,3), activation = "relu", padding = "same") %>%
  layer_max_pooling_2d(c(2,2)) %>%
  layer_conv_2d(4, c(2,2), strides = c(2,1), activation = "relu") %>%
  layer_flatten() %>%
  layer_dense(5, activation = "softmax")

# Convert the model
conv_cnn <- Converter$new(model)
```

### Library `neuralnet`
The usage with nets from the package `neuralnet` is very simple and 
straightforward, because the package offers much less possibilities than 
`torch` or `keras`. The only thing to note is that no custom activation 
function can be used. However, the package saves the names of the inputs and 
outputs, which can of course be overwritten with the arguments 
`input_names` and `output_names` when creating the converter object.

#### Example 1: Dense Network
```{r}
library(neuralnet)
data(iris)

# Create model
model <- neuralnet(Species ~ Petal.Length + Petal.Width, iris, 
                   linear.output = FALSE)

# Convert model
conv_dense <- Converter$new(model)
# Show input names
conv_dense$model_dict$input_names
# Show output names
conv_dense$model_dict$output_names
```


## Step 2: Apply selected Method
The **`innsight`** package provides the following tools for analyzing black box
neural networks based on dense or convolution layers:

- **`Gradient`** : Calculation of the model output **Gradients** with respect
to the model inputs including the attribution method 
[**Gradient x Input**](https://www.jmlr.org/papers/v11/baehrens10a.html):
$$
\begin{align}
\text{Gradient}(x)_i^C &= \frac{\partial f(x)_C}{\partial x_i}\\ 
\text{Gradient x Input}(x)_i^C &= x_i \cdot \text{Gradient}(x)_i^C
\end{align}
$$

- **`SmoothGrad`** : Calculation of the smoothed model output gradients 
([**SmoothGrad**](https://arxiv.org/abs/1706.03825)) with respect to the model
inputs by averaging the gradients over number of inputs with added noise 
(including **SmoothGrad x Input**):
$$
\text{SmoothGrad}(x)_i^C = \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, \sigma^2)}\left[\frac{\partial f(x + \varepsilon)_C}{\partial (x + \varepsilon)_i}\right] \approx \frac{1}{n} \sum_{k=1}^n \frac{\partial f(x + \varepsilon_k)_C}{\partial (x + \varepsilon_k)_i}
$$
with $\varepsilon_1, \ldots \varepsilon_n \sim \mathcal{N}(0, \sigma^2)$.

- **`LRP`** : Back propagation the model output to the model input neurons to 
obtain relevance scores for the model prediction which is known as
[**Layer-wise Relevance Propagation**](https://doi.org/10.1371/journal.pone.0130140):
$$
f(x)_C \approx \sum_{i=1}^d R_i
$$
with $R_i$ relevance score for input neuron $i$.

- **`DeepLift`** : Calculation of a decomposition of the model output with
respect to the model inputs and a reference input which is known as
Deep Learning Important FeaTures 
([**DeepLift**](https://arxiv.org/abs/1704.02685)):
$$
\Delta f(x)_C = f(x)_C - f(x_\text{ref}) = \sum_{i=1}^d C_i
$$
with $C_i$ contribution score for input neuron $i$ to the 
difference-from-reference model output.

- **`ConnectionWeights`** : This is a naive and old approach by calculating 
the product of all weights from an input to an output neuron and then adding
them up (see [**Connection Weights**](https://doi.org/10.1016/j.ecolmodel.2004.03.013)).

#### Notes

* By default, the last activation function is not taken into account 
for all data-based methods . Because often this is the sigmoid/logistic or 
the softmax function, which has with growing distance from 0 
increasingly smaller gradients, which leads to the so-called 
*saturation problem*. But if you still want to consider the last activation 
function, use the argument `ignore_last_act = FALSE`.

* For the data with channels, it is not possible to determine exactly on which 
axis the channels are located. Internally all data and the converted model are 
in the data format 'channels first', i.e. directly after the batch dimension 
`(batch_size, channels, input_dim)`. In case you want to pass data with 
'channels last' (e.g. for MNIST-data *(batch_size, 28, 28, 3)*), 
then you have to indicate that with argument `channels_first` in the
applied method.

* It can happen with very large and deep neural networks that the calculation 
for all outputs requires both the entire memory and takes a very long time. 
But often, the results are needed only for certain output nodes. By default, 
only the results for the first 10 outputs are calculated, which can be adjusted
individually with the argument `output_idx` by passing the relevant output
indices.

### Example 1: Gradient and Gradient x Input
```{r, results='hide', message=FALSE}
# Make sure that Example 1 from the section on neuralnet and one example
# with CNNs were run last

# Apply method 'Gradient' for the dense network
grad_dense <- Gradient$new(conv_dense, iris[-c(1,2,5)])

# Apply method 'Gradient x Input' for CNN
x <- torch_randn(c(10,3,10,10))
grad_cnn <- Gradient$new(conv_cnn, x, times_input = TRUE)
```
### Example 2: SmoothGrad and SmoothGrad x Input
```{r, results='hide', message=FALSE}
# Make sure that Example 1 from the section on neuralnet and one example
# with CNNs were run last

# Apply method 'SmoothGrad' for the dense network
smooth_dense <- SmoothGrad$new(conv_dense, iris[-c(1,2,5)])

# Apply method 'SmoothGrad x Input' for CNN
x <- torch_randn(c(10,3,10,10))
smooth_cnn <- SmoothGrad$new(conv_cnn, x, times_input = TRUE)
```

### Example 3: LRP
```{r, results='hide', message=FALSE}
# Make sure that Example 1 from the section on neuralnet and one example
# with CNNs were run last

# Apply method 'LRP' for the dense network
lrp_dense <- LRP$new(conv_dense, iris[-c(1,2,5)])

# Apply method 'LRP' for CNN with alpha-beta-rule
x <- torch_randn(c(10,10,10,3))
lrp_cnn <- LRP$new(conv_cnn, x, rule_name = "alpha_beta", rule_param = 1,
                   channels_first = FALSE)
```
### Example 4: DeepLift
```{r, results='hide', message=FALSE}
# Make sure that Example 1 from the section on neuralnet and one example
# with CNNs were run last

# Define reference value
x_ref <- array(colMeans(iris[-c(1,2,5)]), dim = c(1,2))
# Apply method 'DeepLift' for the dense network
deeplift_dense <- DeepLift$new(conv_dense, iris[-c(1,2,5)], x_ref = x_ref)

# Apply method 'DeepLift' for CNN
x <- torch_randn(c(10,3,10,10))
deeplift_cnn <- DeepLift$new(conv_cnn, x)
```

## Step 3: Show and Plot the Results
Once a method object has been created, the results can be returned as an 
`array`, `data.frame`, or `torch_tensor`, and can be further processed as 
desired. In addition, for each of the three sizes of the inputs (1D, 2D, or 
3D) suitable plot and boxplot functions based on 
[ggplot2](https://ggplot2.tidyverse.org/) are implemented. Due to the 
complexity of higher dimensional inputs, these plots and boxplots can also 
be displayed as an interactive [plotly](https://plotly.com/r/) plots by using 
the argument `as_plotly`.


### Get Results
Each instance of the interpretability methods have the class method 
`get_result`, which is used to return the results. You can choose between 
the data formats `array`, `data.frame` or `torch_tensor` by passing the
name as an character for argument `type`.

#### Array (Default)
```{r}
# Get result (make sure 'grad_dense' is defined!)
result_array <- grad_dense$get_result()

# Show for datapoint 1 and 71 the result
result_array[c(1,71),,]
```
#### Data Frame
```{r}
# Get result as data.frame (make sure 'grad_cnn' is defined!)
result_data.frame <- lrp_cnn$get_result("data.frame")

# Show the first 5 rows
head(result_data.frame, 5)
```
#### Torch Tensor
```{r}
# Get result (make sure 'deeplift_dense' is defined!)
result_torch <- deeplift_dense$get_result("torch_tensor")

# Show for datapoint 1 and 71 the result
result_torch[c(1,71),,]
```

### Plot Results
The package `innsight` also provides methods for visualizing the results. 
By default a ggplot2-plot is created, but it can also be rendered as an 
interactive plotly plot with the `as_plotly` argument. You can use the 
argument `output_idx` to select the indices of the output nodes 
for the plot. In addition, if the results have channels, the 
`aggr_channels` argument can be used to determine how the channels are 
aggregated.

**Note:** If you want to make a change to the results before plotting, you can 
get the results  with `Method$result` (torch tensor!), change it accordingly, 
and then save it back to the attribute `Method$result` as a torch tensor.

#### Plot Function
```{r}
# Plot result for the first data point and all outputs 
# (make sure 'grad_dense' is defined!)
plot(smooth_dense, output_idx = 1:3)
# You can plot several data points at once
plot(smooth_dense, data_idx = c(1,71), output_idx = 1:3)
# Plot result for the first data point and first and fourth output
# (make sure 'lrp_cnn' is defined!)
plot(lrp_cnn, aggr_channels = 'norm', output_idx = c(1,4))
```
```{r, eval = FALSE}
# Create a plotly plot for the first output
plot(lrp_cnn, aggr_channels = 'norm', output_idx = c(1), as_plotly = TRUE)
```

#### Boxplot Function
```{r}
# Plot result for the first data point and first two outputs 
# (make sure 'smooth_dense' is defined!)
boxplot(smooth_dense, output_idx = 1:2)
# Use no preprocess function (default: abs) and plot reference data point
boxplot(smooth_dense, output_idx = 1:3, preprocess_FUN = identity,
        ref_data_idx = c(55))
```
```{r, eval = FALSE}
# Same as plotly
# In addition, you have a drop-down menu to select other data points
# you can define the indices for the drop-down menu with 'individual_data_idx'
boxplot(smooth_dense, output_idx = 1:3, preprocess_FUN = identity,
        ref_data_idx = c(55), as_plotly = TRUE, individual_data_idx = c(1))
```

